# Optimization

## Introduction

<p> To improve is to make better; to optimize is to make best. Optimization is the act of identifying the extreme (cheapest, tallest, fastest...) over a collection of possibilities. You may recall studying optimization in introductory calculus where you may have solved simple design problems over the possible sizes of fences, or boxes, or something similar. Optimization over __design space__ (also called __decision space__) is a critical feature of many engineering tasks, and has a role in most areas of applied science, including biology. Examples include optimal manipulation of biological systems (e.g. optimal harvesting or optimal drug dosing) or optimal design of biological systems (e.g. robust synthetic genetic circuits). A complementary task is optimal experimental design, which aims to identify the 'best' experiment from a collection of possibilities. Model calibration, to be discussed further below, provides another example; here we seek the 'best fit' version of a model from a collection of possible options.</p>
<p> Beyond its use in design, optimization is also used to investigate natural systems (i.e. in `pure' science), in cases where 'nature' presents us with an optimal version of a given phenomenon. For example, the principle of entropy maximization (equivalently energy minimization) justifies a wide variety of phenomena, from the shapes of soap bubbles to the configuration of proteins.  Darwinian evolution provides another mechanism for optimization. Assuming evolution has arrived at optimal designs, we can apply optimization to understand a variety of biological phenomena, from metabolic flux distribution, to brain wiring, to optimal foraging strategies.</p>
<p> This module introduces optimization methods in R. Applications and illustrations are be drawn from a range of biological domains. The simple optimization tasks addressed by introductory calculus can be accomplished with paper-and-pencil calculations. As shown below, while the fundamental principles introduced by those exercises carry forward, most optimization tasks of interest in biology demand more extensive computational resources.</p>

## Fundamentals of Optimization

Figure 1 illustrates some basics terminology associated with optimization. The graph of a function $f$ of a single variable $x$ is shown, defined over a domain $[a,b]$. In the context of optimization, we can think of each $x$-value in the interval $[a,b]$ as one possible scenario (e.g. fence length, harvesting rate, etc.). The function $f$ maps those choices to some __objective__ (e.g. total area, long-term annual yield) that we wish to optimize (either maximize or minimize). The _global_ extrema (i.e. maximum and minimum) represent the results we wish to achieve.  

<p><center>
`r htmltools::img(src=knitr::image_uri("Figure1.png"))`  
Figure 1: Extreme Values  
</center></p>

<p> More generally, we define _local_ extrema (maxima and minima) as cases that appear to be optimal if we restrict our attention only to 'nearby' possibilities (i.e. $x$-values). There is an extensive theory of optimization methods; unfortunately most of these are dedicated to __identifying local optima__. These approaches cannot directly identify global optimum -- instead they identify _candidate_ global optima, which then can be compared to identify the 'best' result. (A convenient special case occurs for problems where every local optimum is a global optimum; these are called convex optimization problems. Unfortunately, they occur only rarely in addressing biological phenomena.) </p>

### Fermat's Theorem

<p> To illustrate these fundamentals, consider the following two academic examples that rely on basic calculus, specifically on __Fermat's Theorem__, which states that local extrema occur at points where the tangent line to a function's graph is horizontal, i.e. at points where the derivative (i.e. the slope of the tangent) is zero (Figure 2). </p>

<p><center>
`r htmltools::img(src=knitr::image_uri("Figure2.png"))`  
Figure 2: Fermat's Theorem: local extrema occur at points where the tangent line is horizontal. Some local extrema are also global extrema.
</center></p>


<p> Example 1.  Identify the value of $x$ for which the function $f(x) = x^2+3x-2$ is minimized. </p>
<p> Solution: Taking the derivative, we find $f'(x) = 2x+3$. To find the point(s) where the derivative is zero, we solve: $f'(x) = 0 \Leftrightarrow 2x+3 = 0\Leftrightarrow x = -\frac{3}{2} = -1.5$. As shown in Figure 4, the single local minimum is the global miminum in this case, so $x=-1.5$ is the desired solution.</p> 


<p><center>
`r htmltools::img(src=knitr::image_uri("Figure3.png"))`  
Figure 3: Graph of $f(x) = x^2+3x-2$  (Example 1)   
</center></p>

<p> Example 2.  Identify the value of $x$ at which the function $f(x) = 3x^4-4x^3-54x^2+108x$ is minimized. </p>
<p> Solution: Taking the derivative, we find $f'(x) = 12x^3-12x^2-108x+108 = 12(x-3)(x-1)(x+3)$. To find the point(s) where the derivative is zero, we solve: $f'(x) = 0 \Leftrightarrow x = 3, 1, -3$. As shown in Figure 4, two of these points are local minima. One ($x=-3$) is where the global minimum occurs.</p> 


<p><center>
`r htmltools::img(src=knitr::image_uri("Figure4.png"))`  
Figure 4: Graph of $f(x) = 3x^4-4x^3-54x^2+108x$  (Example 2)
</center></p>


## Regression

### Linear Regression

<p> As mentioned above, finding the 'best fit' from a family of models is a common optimization task in science. The simplest example of this task is __linear regression__: determining a line of best fit through a given set of points. </p> 
<p> To illustrate, consider the dataset of $(x,y)$ pairs shown in Figure __XYZ__, which we can label as $(x_1, y_1)$, $(x_2, y_2)$, ldots, $(x_N, y_N)$. Several lines are displayed. The optimization task here is to identify the 'best' line: the one that best captures the trend in the data. To specify this task  mathematically, we need to decide on a measure of 'quality of fit'. We start by recognizing that the line will (typically) fail to pass through most of the points in the dataset. Thus, at each point $x_i$ we can define an __error__, which is the difference between the observed $y$-value and the $y$-value on the line. If we call the line $y=mx+b$, then the error at $x_i$ can be defined as $e_i = y_i-(mx_i+b)$. We then need to combine these errors into a single measure of teh quality of fit. This is typically done by _squaring_ the errors and adding them together. (Squaring ensures that both under- and over-estimations contribute equally.) We define the __sum of squared errors__ (SSE) as : $(y_1-(mx_1+b))^2 + (y_2-(mx_2+b))^2 + \cdots +(y_N-(mx_N+b))^2 = e_1^2+e_2^2+ \cdots e_N^2$. </p>

<p> We can now pose the model-fitting task as an optimization problem. For each line (that is, each assignment of numerical values to $m$ and $b$), we associate a corresponding SSE. We seek the values of $m$ and $b$ for which the SSE is a global minimum. </p>


<p> Example 3. Consider a simplified version of linear regression, in which we know that our model (line) should pass through the origin (0,0). That is, instead of lines $y=mx+b$, we will consider only lines of the form $y=mx$. (We thus have a single parameter to identify: the slope $m$.) To keep the algebra as simple as possible we'll take a tiny dataset consisting of just two points: $(x_1, y_1) = (2,3)$ and $(x_2, y_2)= (5.4)$, as indicated in Figure 5. The line passes through points $(2,2m)$ and $(5, 5m)$. </p>

<p><center>
`r htmltools::img(src=knitr::image_uri("Figure5.png"))`  
Figure 5: Example 1: finding a line of best fit through the origin  
</center></p>


In this simple case, the sum of squared errors is:
\begin{equation*}
    \mbox{SSE} = e_1^2+e_2^2 = (3-2m)^2+(4-5m)^2
\end{equation*}  
To apply Fermat's theorem we take the derivative and identify any values of $m$ for which it is zero:
\begin{eqnarray*}
    \frac{d}{dm} \mbox{SSE} &=& 2(3-2m)(-2)+2(4-5m)(-5) \\
    &=& -4(3-2m) - 10(4-5m) = -12 +8m+-40+50m = -52+58m.
\end{eqnarray*}
<p> The only local extremum (and hence the only candidate for global minimum) is then $m=52/58 = 26/29$.</p>
<p> The analysis in Example 3 can be extended to determine the general solution of the linear regression task [Fairway, 2002]. The solution formula is as follows:
__Linear regression formula__ The best fit line $y=mx+b$ to the dataset $(x_1, y_1)$, $(x_2, y_2)$, \ldots, $(x_n, y_n)$ is given by
\begin{eqnarray*}
    m&=& \frac{\sum_{i=1}^n (x_i-\bar{x})(y_i-\bar{y})}{\sum_{i=1}^n (x_i-\bar{x})^2} \\
    b&=& \bar{y}-m\bar{x}, \\
\end{eqnarray*}
where $\bar{x}$ and $\bar{y}$ are the averages
\begin{eqnarray*}
    \bar{x}=\frac{1}{n} \sum_{i=1}^n x_i \ \ \ 
\bar{y}=\frac{1}{n} \sum_{i=1}^n y_i
\end{eqnarray*}  </p>

<p> This formula is somewhat unwieldy, but is straightforward to implement. In R, the command ```lm``` implements this formula, as the following example illustrates. </p>

We'll work with a dataset called __cars__ that is included in R. It's shown below (via the __plot()__ command).


```{r, echo = TRUE}
#plot(cars)
```

The __lm()__ command applies the formula described above to arrive at the line of best fit (i.e. the minimizer of the sum of squared errors.  We use teh command in the form  __lm(dist~speed)__, referring to the labels for the data (as shown above) to generate the parameters ($b$ and $m$) that specify the line $y=mx+b$. (The __attach__ command ensures that the data is accessible.)

```{r, echo = TRUE}
#attach(cars)
#lm(dist~speed)
```

The intercept is -17.579 and the slope is 3.932. 

We can plot the data together with the line of best fit with the __abline__ command: 
<!-- or __abline(lm(dist~speed))__.  -->


```{r, echo = TRUE}
#attach(cars)
#plot(cars)
#abline(-17.579, 3.932)
```

### Nonlinear Regression

<p> Nonlinear regression is the task of fitting a nonlinear model (e.g. a curve) to a dataset. The setup for this task is identical to linear regression. We begin by selecting a parameterized family of models (i.e. curves), and aim to identify the model (i.e. curve) that minimizes the sum of squared errors when compared to the data. The family of models can be chosen based on an understanding of the mechanism that relates the input and output data. For example, if we are investigating  the rate law for a single-substrate enzyme-catalysed reaction, we might choose the family of curves specified by Michaelis-Menten kinetics: 
\begin{eqnarray*}
V = \frac{V_{\mbox{max}} S}{K_M+S}
\end{eqnarray*} 
Our goal would then be to identify values for the parameters $V_{\mbox{max}}$ and $K_M$ that minimize the sum of squared errors when comparing with observed data. </p>
<p> __Regression via linearizing transformation__
In several important cases, the nonlinear regression task can be transformed into a linear regression task. In the case of Michaelis-Menten kinetics, several linearizing transformations have been proposed (e.g. Eadie–Hofstee and Lineweaver–Burk [Cho and Lim, 2018]). Another example commonly encountered in Biology is fitting exponential curves (e.g. population growth or drug clearance).  In those cases, a logarithm modifies the data so that a linear trend is captured: e.g. $y = e^{rt}$, after applying a logarithm becomes $y^{trans} = \ln(y) = \ln (e^{rt})= rt$. Linear regression on transformed data $(t_i, y^{trans}_i)$ then provides an estimate of the value of $r$.  </p>

<p> Unfortunately, linearizing transformations are only available in a handful of special cases. Moreover, they often introduce biases that can make interpretation of the resulting model difficult. 

In general, the nonlinear regression task must be addressed directly. An example of the general procedure in R follows. We begin by defining a dataset against which we will fit a Michaelis-Menten curve. 

```{r, echo = TRUE}
MM_data_old <- structure(list(S = c(4.32, 2.16, 1.08, 0.54, 0.27, 0.135, 3.6, 
                           1.8, 0.9, 0.45, 0.225, 0.1125, 2.88, 1.44, 0.72, 0.36, 0.18, 0.9, 
                           0), v = c(0.004835714, 0.004192308, 0.003553846, 0.002576923, 
                                     0.001661538, 0.001064286, 0.004407692, 0.004671429, 0.0039, 0.002857143, 
                                     0.00175, 0.001057143, 0.004907143, 0.004521429, 0.00375, 0.002764286, 
                                     0.001857143, 0.001121429, 0)), .Names = c("S", "V"), class = "data.frame",
                row.names = c(NA,-19L))

S <- c(4.32, 2.16, 1.08, 0.54, 0.27, 0.135, 3.6, 
                           1.8, 0.9, 0.45, 0.225, 0.1125, 2.88, 1.44, 0.72, 0.36, 0.18, 0.9, 
                           0) 
V <- c(0.004835714, 0.004192308, 0.003553846, 0.002576923, 
                                     0.001661538, 0.001064286, 0.004407692, 0.004671429, 0.0039, 0.002857143, 
                                     0.00175, 0.001057143, 0.004907143, 0.004521429, 0.00375, 0.002764286, 
                                     0.001857143, 0.001121429, 0)
MM_data=cbind(S,V)
plot(MM_data)
```

Like the ```lm``` command, the ```nls``` function takes as inputs the dataset and the model. In addition, ```nls``` requires that the user provides 'guesses' for the values of the parameters to be estimated (in this case $V_{\mbox{max}}$ and $K_M$). In this case, we can roughly estimate $V_{\mbox{max}} \approx  0.005$ as the maximal $V$-value that can be achieved, and $K_M \approx 0.2$ as the $S$-value at which $V$ reaches half of its maximal value.

```{r, echo = TRUE}
model.nls <- nls(V ~ Vmax * S/(Km+S),start = list(Km = 0.3, Vmax = 0.005))
summary(model.nls)
```

This output from ```nls``` provides the parameter estimates that characterize the best-fit model ($K_M = 0.4772154$ and $V_{\mbox max} = 0.0053061$), as well as additional information about the confidence of this estimate (some of which will be discussed further below). We can now plot the best fit model along with the dataset:  

```{r, echo = TRUE}
params <- summary(model.nls)$coeff[,1] #extracting the parameter estimates
plot(MM_data)
curve((params[2]*x)/(params[1]+x), from = 0, to = 4, add=TRUE, col='firebrick')
```
</p>
<p> From the implementation of ```nls```, the reader might have the mistaken impression that nonlinear regression and linear regression are very similar tasks.  Although the problem set-up is similar (first chose family of curves, then minimize SSE), the optimization task is very different, and the solution procedures follow very different protocols. (The first hint of this is the need to provide ``guesses'' to ```nls```.) As we saw above, the solution to the linear regression task can be stated as a general formula. For nonlinear regression, no such formula exists. Worse, as we'll discuss in the next section there is no procedure (algorithm) that is guaranteed to find the solution! </p>

## Iterative Optimization Algorithms

<p> The best techniques for addressing the general nonlinear regression task are iterative global optimization routines. As we'll discuss below, these algorithms all follow a basic idea: we start with a 'guess' which established initial estimates for the parameter values and then take steps through parameter space to improve the quality of that estimate. In the exercise above, the ```nls``` command executed this kind of algorithm, which is why it requires that the user supply an 'initial guess'. </p>

### Gradient Descent

<p> A simple iterative optimization algorithm is __gradient descent__, which can be understood intuitively via a thought experiment. Imagine finding your way to a valley bottom in a think fog. The fog obscures your vision so that you can only detect changes in elevation in your immediate vicinity. To make your way to the valley bottom, it would be reasonable to take each step of your journey in the direction of steepest decline. This strategy is guaranteed to lead to a local minimum, but cannot guarantee arrival at the lowest point: the global minimum. That this, the search may lead to a local minimum at which there is no direction of local descent, and so the algorithm gets `stuck'.</p>

<p> Mathematically, the local change in elevation is determined by evaluating the optimization objective at points nearby the current estimate, and then using those to determine the direction of steepest descent. (technically, this involves a linearization of the function at the current position, or equivalently a calculation of the _gradient vector_). A step is then taken in the direction, and the process is repeated from this updated estimate. To implement this algorithm, a number of details have to be specified: how long should each step be? How many steps should be taken? Or should there be some other 'termination condition' that will trigger the end of the journey? (Each of these decisions involve a trade-off, typically a trade-off of precision vs. computation time. For instance, taking very small steps will guarantee a smooth path down the steepest route, but might take a very long time to complete the journey. Termination conditions are often specified in terms of the local topography: the algorithm stops when the current estimate is at a sufficiently flat position (no downhill direction detected). </p>

<p>To illustrate the gradient descent approach, consider the following algorithm, which incorporates two termination conditions: a maximum number of allowed iterations,and a threshold for shallowness: </p>

```{r,echo=TRUE}
library(numDeriv)       # contains functions for calculating gradient

#define function that implement gradient descent. Inputs are the objective function f, the initial parameter estimate x0, the maximum number of iterations to be applied, the size of each step, and a threshold gradient below which the landscape is considered flat (and so iterations are terminated)
grad.descent = function(f, x0, step.size=0.05, max.iter=200, 
                        stopping.deriv=0.01, ...) {
  n = length(x0)  # record the number of parameters to be estimated (i.e. the dimension of the parameter space)
  xmat = matrix(0,nrow=n,ncol=max.iter) # initialize a matrix to record the sequence of estimates
  xmat[,1] = x0   # the first row of matrxi xmat is the initial estimate 'guess'
  for (k in 2:max.iter) {
     # Calculate the gradient (a vector indicating steepness and direct of greatest ascent)
    grad.current = grad(f,xmat[,k-1],...)   
    # Check termination condition: has a flat valley bottom been reached?
    if (all(abs(grad.current) < stopping.deriv)) {
      k = k-1; break
      3
    }
    # Step in the opposite direction of the grad
    xmat[,k] = xmat[,k-1] - step.size * grad.current
  }
  xmat = xmat[,1:k] # Trim any unused columns from xmat
  return(list(x=xmat[,k], xmat=xmat, k=k))
}
```

####Example 1
 <p>We'll begin by demonstrating the performance of this algorithm on a simple function with a single local minimum. </p>

```{r}
simpleFun = function(x) {
  return(x[1]^2 + 1/3*x[2]^2)
}
```

```{r,echo=FALSE,message=FALSE}
simpleFun = function(x) {
  return(x[1]^2 + 1/3*x[2]^2)
}

simpleplotFun = function(x,y) {
  return(x^2 + 1/3*y^2)
}
x <- seq(-2, 2, length= 30)
y <- x
z <- outer(x, y, simpleplotFun)
z[is.na(z)] <- 1

require(lattice)
wireframe(z, drape=T, expand = 0.5, col = "blue", xlab=NULL, ylab=NULL, zlab=NULL,
          scales = list(arrows = FALSE))

```

 <p>In this case, we expect that the minimum (valley bottom) will be reached from any initial 'guess'. Starting from four distinct points, the algorithm follows the paths shown below. </p>

```{r,echo=FALSE}
x0 = c(-1.9,-1.9)
gd = grad.descent(simpleFun,x0,step.size=1/3)

x1 = c(-1.9,1.1)
gd1 = grad.descent(simpleFun,x1,step.size=1/3)

x2 = c(1.9,-1.9)
gd2 = grad.descent(simpleFun,x2,step.size=1/3)

x3 = c(1.9,1.1)
gd3 = grad.descent(simpleFun,x3,step.size=1/3)

surface = function(f, from.x=0, to.x=1, from.y=0, to.y=1,
                   n.x=30, n.y=30, theta=5, phi=80, ...) {
  # Build the 2d grid
  x.seq = seq(from=from.x,to=to.x,length.out=n.x)
  y.seq = seq(from=from.y,to=to.y,length.out=n.y)
  plot.grid = expand.grid(x.seq,y.seq)
  z.vals = apply(plot.grid,1,f)
  z.mat = matrix(z.vals,nrow=n.x)
  # Plot with the persp function
  orig.mar = par()$mar # Save the original margins
  par(mar=c(1,1,1,1)) # Make the margins small
  r = persp(x.seq,y.seq,z.mat,theta=theta,phi=phi,...)
  par(mar=orig.mar) # Restore the original margins
  invisible(r)
}
```



```{r,echo=FALSE}
r = surface(simpleFun,from.x=-2,to.x=2,from.y=-2,to.y=2,xlab="",ylab="",zlab="")
points(trans3d(x0[1],x0[2],simpleFun(x0),r),col="red",cex=2)
points(trans3d(gd$xmat[1,],gd$xmat[2,],apply(gd$xmat,2,simpleFun),r),
      col="blue")
lines(trans3d(gd$xmat[1,],gd$xmat[2,],apply(gd$xmat,2,simpleFun),r),
      lwd=2,col="green")
points(points(trans3d(gd$x[1],gd$x[2],simpleFun(gd$x),r),
              col="black",cex=2))

points(trans3d(x1[1],x1[2],simpleFun(x1),r),col="red",cex=2)
points(trans3d(gd1$xmat[1,],gd1$xmat[2,],apply(gd1$xmat,2,simpleFun),r),
      col="blue")
lines(trans3d(gd1$xmat[1,],gd1$xmat[2,],apply(gd1$xmat,2,simpleFun),r),
      lwd=2,col="orange")
points(points(trans3d(gd1$x[1],gd1$x[2],simpleFun(gd1$x),r),
              col="black",cex=2))

points(trans3d(x2[1],x2[2],simpleFun(x2),r),col="red",cex=2)
points(trans3d(gd2$xmat[1,],gd2$xmat[2,],apply(gd2$xmat,2,simpleFun),r),
      col="blue")
lines(trans3d(gd2$xmat[1,],gd2$xmat[2,],apply(gd2$xmat,2,simpleFun),r),
      lwd=2,col="firebrick")
points(points(trans3d(gd2$x[1],gd2$x[2],simpleFun(gd2$x),r),
              col="black",cex=2))

points(trans3d(x3[1],x3[2],simpleFun(x3),r),col="red",cex=2)
points(trans3d(gd3$xmat[1,],gd3$xmat[2,],apply(gd3$xmat,2,simpleFun),r),
      col="blue")
lines(trans3d(gd3$xmat[1,],gd3$xmat[2,],apply(gd3$xmat,2,simpleFun),r),
      lwd=2,col="magenta")
points(points(trans3d(gd3$x[1],gd3$x[2],simpleFun(gd3$x),r),
              col="black",cex=2))
```

 <p>The global minimum was reached in every case. </p>

####Example 2

 <p>Next, let's consider a  function that has multiple local minima. The second plot below is interactive, allowing you to rotate the surface.</p>   

```{r}
complicatedFun = function(x) {
  return((1/2*x[1]^2-1/4*x[2]^2+3)*cos(2*x[1]+1-exp(x[2])))
}
```
```{r,echo=FALSE,message=FALSE}
complicated = function(x,y) {
  return((1/2*x^2-1/4*y^2+3)*cos(2*x+1-exp(y)))
}
x <- seq(-2, 2, length= 30)
y <- x
z <- outer(x, y, complicated)
z[is.na(z)] <- 1

require(lattice)
wireframe(z, drape=T, expand = 0.5, col = "blue", xlab=NULL, ylab=NULL, zlab=NULL,
          scales = list(arrows = FALSE))
```


```{r,echo=FALSE,message=FALSE,warning=FALSE}
library(MASS)
library(plotly)
fig <- plot_ly(x = x, y = y, z = z,color = "blue") %>% add_surface()

fig
```

<p> Again, we'll apply gradient descent from a set of initial guess positions: </p>

```{r,echo=FALSE}
x0 = c(0.5,0.5)
gd0 = grad.descent(complicatedFun,x0,step.size=0.01)
x1 = c(-0.1,-1.3)
gd1 = grad.descent(complicatedFun,x1,step.size=0.01,max.iter=400)
x2 = c(-1.5,1.3)
gd2 = grad.descent(complicatedFun,x2,step.size=0.01,max.iter=400)
x3 = c(-0.2,1.4)
gd3 = grad.descent(complicatedFun,x3,step.size=0.01,max.iter=400)
x4 = c(-0.5,-0.5)
gd4 = grad.descent(complicatedFun,x4,step.size=0.01,max.iter=400)

r = surface(complicatedFun,from.x=-2,to.x=2,from.y=-2,to.y=2,
xlab="",ylab="",zlab="")
lines(trans3d(gd0$xmat[1,],gd0$xmat[2,],apply(gd0$xmat,2,complicatedFun),r),
      lwd=2,col="red")
points(trans3d(gd0$xmat[1,],gd0$xmat[2,],apply(gd0$xmat,2,complicatedFun),r),
      col="red")
points(trans3d(gd4$xmat[1,],gd4$xmat[2,],apply(gd4$xmat,2,complicatedFun),r),
      col="blue")
lines(trans3d(gd4$xmat[1,],gd4$xmat[2,],apply(gd4$xmat,2,complicatedFun),r),
      lwd=2,col="blue")
lines(trans3d(gd2$xmat[1,],gd2$xmat[2,],apply(gd2$xmat,2,complicatedFun),r),
      lwd=2,col="purple")
points(trans3d(gd2$xmat[1,],gd2$xmat[2,],apply(gd2$xmat,2,complicatedFun),r),
      col="purple")
lines(trans3d(gd3$xmat[1,],gd3$xmat[2,],apply(gd3$xmat,2,complicatedFun),r),
      lwd=2,col="darkgreen")
points(trans3d(gd3$xmat[1,],gd3$xmat[2,],apply(gd3$xmat,2,complicatedFun),r),
      col="darkgreen")
lines(trans3d(gd1$xmat[1,],gd1$xmat[2,],apply(gd1$xmat,2,complicatedFun),r),
      lwd=2,col="orange")
points(trans3d(gd1$xmat[1,],gd1$xmat[2,],apply(gd1$xmat,2,complicatedFun),r),
      col="orange")
```

<p> Here we see that from some initial guess values, the algorithm successfully reaches the global minimuml from others, it gets stuck at a local minimum. </p>


<p> The default optimization algorithm used by ```nls``` is the Gauss-Newton method, which is a generalization of Newton's method for solving nonlinear equations (which may be familiar from introductory calculus). This is a refinement of gradient descent in which the local curvature of the function is used to project the position of the bottom of the local valley _assuming that the surface is shaped like a parabola_. On parabolic surfaces, this achieves descent to the bottom in one iteration. On non-parabolic surfaces (i.e.~most surfaces), the results is a sequence of iterations, but often many fewer than would be required using gradient descent. </p>

<p> Both gradient descent and Gauss-Newton method are design to reach the bottom of the valley in which the initial guess lies. If that's  not the global minimum, then the algorithm will not be successful.  So, how does one select a 'good' initial guess? Unfortunately, there's no general answer to this question. In many cases, one can use previous knowledge of the system under investigation to begin with a solid initial guess. If no such previous knowledge is available, sometimes a `guess' is all we have. In those cases, we may have little confidence that the algorithm will arrive at a global minimum. The simplest way to gain some confidence of achieving a global minimum is the multi-start strategy: choose many initial guesses, and run the algorithm from each (as in the example above). This can be computationally expensive, but if the initial guesses are spread widely over the parameters space, one can have hope that the global minimum will be achieved. </p>

<p> A number of methods have been developed to complement the multi-start approach. These are known as __global optimization routines__. They are also known as _heuristic methods_, because their performance cannot be guaranteed in general: there are no guarantees that they'll find the global minimum, nor are there solid estimates of how many iterations will be required for them to carry out a satisfactory search of the parameter space. We will consider two commonly used heuristic methods: simulated annealing and genetic algorithms. </p>

#### Simulated Annealing

<p> __Simulated annealing__ is motivated by the process of annealing: a heat treatment by which the physical properties of metals can be altered. Simulated annealing is an iterative algorithm; the algorithm starts at an initial guess, and then steps through the search space. In contrast with gradient descent, there is a random element to the process followed by simulated annealing. Consequently, the path followed from a particular initial condition won't be repeated if the algorithm is run again from the same point. (Algorithms that incorporate randomness are often referred to as _Monte Carlo_ methods, in reference to the European gambling centre.)  At each step, the algorithm begins by identify a candidate next position. (This point could be selected by a variant of gradient descent, or some other method).  The value of the objective at this candidate point is then compared with the objective value at the current point. If the objective is lower at the new point (i.e. this step takes the algorithm downhill), then the step is taken and a new iteration begins. If the value of the objective is larger at the candidate (i.e. the step makes things _worse_), the step can still be taken, but only with a small probability. Both the size of the candidate steps and the probability of accepting 'wrong' (uphill) steps are tied to a __cooling schedule__: a decreasing `temperature' profile: at high temperatures, large steps are considered and 'wrong' steps are taken frequently; as the temperature drops, only smaller steps are considered, and fewer 'wrong' steps are allowed. By analogy, imagine a ping-pong ball resting on a curved landscape. One strategy to move the ball to the lowest valley bottom is to shake the table. Mirroring simulated annealing, we could begin by applying violent shakes (high temperature) which would result in the ball bouncing across much of the landscape. By slowly reducing the severity of the shaking, the ball would settle into a local minimum at a valley bottom. The hope is that if the cooling schedule is well-chosen, the ball would have sampled many valleys, and would end up at the bottom of the lowest valley. Simulated annealing is often combined with a multi-start strategy to further ensure broad sampling of the search space. </p>

<p> To illustrate, we'll apply simulated annealing to the optimization task in Example 2 above, using the same initial guess points. We'll use the ```GenSA``` library to implement the algorithm (add link https://cran.r-project.org/web/packages/GenSA/GenSA.pdf). Calls to ```GenSA``` requires that we specify the objective function, an initial guess, and upper and lower bounds for the search values for each parameter. (Optional input parameters allow the user to modify internal features of the algorithm such as the cooling schedule and stepping protocol.)

```{r,echo=TRUE}
library(GenSA)

out0 <- GenSA(par = x0, lower = c(-2,-2), upper = c(2,2),fn = complicatedFun)
out0[c("value","par")]
```
<p> The result of this call is recorded in the variable ```out0``` which indicates the minimal value of the objective achieved (-4.906786) and the parameter values at which this minimum occurs $(x,y)= (2.0000000, 0.6014984)$. Next, we'll call ```GenSA``` form each of the initial guesses that were provided to the gradient descent algorithm above. </p>

```{r,message=FALSE}
library(GenSA)

set.seed(1234) # The user can use any seed.
dimension <- 2
global.min <- 0
tol <- 1e-13
lower <- rep(-2, dimension)
upper <- rep(2, dimension)

out0 <- GenSA(par = x0, lower = lower, upper = upper,fn = complicatedFun)
out0[c("value","par","counts")]

out1 <- GenSA(par = x1, lower = lower, upper = upper,fn = complicatedFun,control = list(temperature=2000))
out1[c("value","par","counts")]

out2 <- GenSA(par = x2, lower = lower, upper = upper,fn = complicatedFun)
out2[c("value","par","counts")]

out3 <- GenSA(par = x3, lower = lower, upper = upper,fn = complicatedFun)
out3[c("value","par","counts")]

out4 <- GenSA(par = x4, lower = lower, upper = upper,fn = complicatedFun)
out4[c("value","par","counts")]
```

<p> We see that simulated anneaing has avoided getting 'stuck' in the local minima, and has achieved the same optimized value from every initial guess. </p>

<p> The plot below provides some insight into how the simulated annealing run proceeded. Iterations (steps) are shown along the horizontal axis. The vertical axis shows values of the objective function. The blue points represents function value at teh current position, while the red shows the minimum achieved so far. The minimum is found rather quickly, but the algorithm continues to jump around (and occasionally get temporarily stuck at local minima) in an attempt to widely explore the search space.</p>

```{r}
plot(out1$trace.mat[,4][1:8000],type="l",lwd=4,col="red")
points(out1$trace.mat[,3][1:8000],pch=20,col="blue")
```

####Example 4
<p> To give the simulated annealing algorithm a more challeneging task, consider the following function, which has many local minima. </p>

```{r,echo=FALSE}
f.name <- "Eggholder function"

# make the function
f.egg<- function(x){
  -(x[,2]+47)*sin(sqrt(abs(x[,2]+(x[,1]/2)+47)))-x[,1]*sin(sqrt(abs(x[,1]-(x[,2]+47))))
}

# define the function
f.egghold <- function(x) {
  # make x a matrix so this function works for plotting and for optimizing
  x <- matrix(x, ncol=2)
  # calculate the function value for each row of x
  f.x <- f.egg(x)
  # return function value
  return(f.x)
}

# plot the function
# define ranges of x to plot over and put into matrix
x1 <- seq(-512, 512, length = 101)
x2 <- seq(-512, 512, length = 101)
X <- as.matrix(expand.grid(x1, x2))
colnames(X) <- c("x1", "x2")

# evaluate function
y <- f.egghold(X)

# put X and y values in a data.frame for plotting
df <- data.frame(X, y)
# plot the function
library(lattice) # use the lattice package
wireframe(y ~ x1 * x2 # y, x1, and x2 axes to plot
  , data = df # data.frame with values to plot
  , main = f.name # name the plot
  , shade = TRUE
  , scales = list(arrows = FALSE) # include axis ticks
  , screen = list(z = -100, x = -30) # view position
)
```

```{r,echo=FALSE}
f.egg<- function(x){
  -(x[2]+47)*sin(sqrt(abs(x[2]+(x[1]/2)+47)))-x[1]*sin(sqrt(abs(x[1]-(x[2]+47))))
}

lower <- rep(-200, dimension)
upper <- rep(200, dimension)

x0 <- c(100,100)
```

Here we started from two different starting points and they resulted in the same final position.

```{r}
egg.out <- GenSA(par = x0, lower = lower, upper = upper,fn = f.egg)
egg.out[c("value","par","counts")]

egg.out3 <- GenSA(par = x3, lower = lower, upper = upper,fn = f.egg)
egg.out3[c("value","par","counts")]
```

The plot below shows how simulated annealing works for the two initial points respectively. The red line records the current minimum. The blue points represents function value. Although they have different starting position, it can be seen from the plots the optimized values are the same.

```{r,echo=FALSE}
plot(egg.out$trace.mat[,4],type="l",lwd=4,col="red")
points(egg.out$trace.mat[,3],pch=20,col="blue")

plot(egg.out3$trace.mat[,4],type="l",lwd=4,col="red")
points(egg.out3$trace.mat[,3],pch=20,col="blue")
```

We next consider a heuristic algorithm in which multiple paths through the search space are followed simultaneously.  

#### Genetic Algorithms

<p> __Genetic algorithms__ are inspired by Darwinian evolution. The algorithm begins with the specification of a population of initial guesses. At each iteration of the algorithm, this population 'evolves' toward improved estimates of the global minimum. This 'evolution' step involves three substeps: selection, mutation, and cross-over. In the selection step, the population is pruned by removing a fraction that are not sufficiently fit (where fitness corresponds to the value of the objective function being minimized). Then, mutations are introduced into the remaining population by adding small random perturbations to their position in the search space. Finally, a new generation is generated by crossing members of the current population. This can be done in several ways; the simplest is to generate crosses as averages of the numerical values of the two 'parents'. The hope is that, through several generations, this process will lead to a population with high fitness (minimal objective) after a thorough exploration of the search space. Genetic algorithms are a subset of the more general class of _evolutionary algorithms_ all of which involve simultaneous 'exploration' of the search space through multiple paths.  </p>

</p>To implement a genetic algorithm, we'll make use of the ```ga``` function (https://cran.r-project.org/web/packages/GA/GA.pdf). As with simulated annealing, the call to ```ga``` requires that we specify the objective function, lower and upper bounds to define the search space, and the number of iterations to execute. B default, ```ga``` maximizes the objective function, so we enter our objective with a negative sign to achieve minimization. </p>

```{r}
library(GA)
ga <- ga(type = "real-valued", fitness = function(x) -complicated(x[1],x[2]),
         lower = c(-2, -2), upper = c(2, 2), maxiter = 30)
```

Plotting the results of the genetic algorithm search, we see improvement in the overall behaviour from generation to generation. 

```{r}
plot(ga)
```

The summary displayed below shows that the solutions reached by the genetic algorithm agrees with the soltuion found by simulated annealing.  

```{r}
summary(ga)
```

Example: maybe biexponential or not, depending on complexity of cost surface. </p>

## Calibration of Dynamic Models

<p> The principles of nonlinear regression carry over directly to calibtration of more complex models. In particular, in many domains of biology, dynamic models are used to describe the time-varying behaviour of systems (from biomolecular networks to cell-cell interactions to physiology to ecology). These models take many forms, but a commonly used formulation is a model based on ordinary differential equations (i.e. _rate equations_). These models are deterministic (cannot incorporate random effects) and assume that the dynamics occur in a spatially homogeneous environment (they capture spatially distributed phenomena). Despite these limitations, these models can describe a wide variety of dynamic behaviours, and so are useful starting points for investigations across biology. </p>
<p> Ordinary differential equation models used in biology often take the form
\begin{equation*}
    \frac{d}{dt} {\bf x}(t) = {\bf f}({\bf x}(t), {\bf p})
\end{equation*}
where components of the time-varying vector ${\bf x}(t)$ are the _states_ of the system (e.g.~population sizes, molecular concentrations), the components of vector ${\bf p}$ are the _model parameters_: numerical values that represent fixed features of the system and its environment (e.g. interaction strengths, temperature, nutrient availability), and the vector-valued function ${\bf f}$ describes the rate of change of the state variables. As a concrete example, consider the Lotka-Volterra equations, a classical model to describe interacting predator and prey populations [ref]:
\begin{eqnarray*}
    \frac{d}{dt} x_1(t) &=& p_1 x_1(t) - p_2 x_1(t) x_2(t)\\
    \frac{d}{dt} x_2(t) &=& p_3 x_1(t) x_2(t) - p_4 x_2(t)\\
\end{eqnarray*} </p>
Here $x_1$ is the size of the prey population; $x_2$ is the size of the predator population. The prey are presumed to have access to resources that allow exponential growth in the absence of predation (growth at rate $p_1 x_1(t)$. Interactions between prey and predator populations (assumed to occur at rate $x_1(t) x_2(t)$ lead to an decrease in the prey population and an increase in the predator population (characterized by parameters $p_2$ and $p_3$ respectively). Finally, the prey suffer an exponential decline in population size in the absence of prey (decay rate $p_4 x_2(t)$). A simulation of the model is shown in Figure. Note that simualtion of the model requires specification of (i) values for each of the model parameters, and (ii) initial conditions, i.e.~the size of each population at time $t=0$.

<p> Figure XYZ shows a dataset the corresponds to observations of a predator-prey population system. To calibrate the model Lotka-Volterra to this dataset we seek values for the four parameters $p_1$, $p_2$, $p_3$, and $p_4$ for which simulations of the model provide the 'best fit'. As in the regression tasks described previously, the standard measure for quality of fit is the sum of squared errors. We thus proceed with a minimization task: for each simulation of the model, we compare with the dataset and determine the sum of squared errors. We aim to minimize this fit over the space of model parameters. There's one additional feature we must account for: to specify a simulation we need numerical values for the model parameters and the initial conditions. Calling these initial conditions $p_5 = x_1(0)$ and $p_6=x_2(0)$, we are then pose our optimization problem as a search over a six-dimensional parameter space. In what follows, we illustrate the use of both simulated annealing and a genetic algorithm to calibrate this model.</p>

<p> Matt: insert example </p>

<!-- ## Other Topics -->

<!-- ### Optimal Control -->

<!-- <p> The behaviour of the Lotka-Volterra model described above depends on the values of the model parameters. Such models are often used to explore the effect of perturbations on a system. For instance, after successfully calibrating a model to given population's dynamics, one could propose interventions that could alter the value of one of the model parameters, e.g.~by restricting access to resources required for growth of the prey population. We could then simulate the model under this altered value of $p_1$ and thus generate predictions of the effect of this manipulation. </p> -->

<!-- <p> Alternatively, we could use a model formulation that incorporates a time-varying perturbation.  As an example, consider a version of the Lotka-Volterra equations that includes a term to describe removal (harvesting/culling) of the predator population: -->
<!-- \begin{eqnarray*} -->
<!--     \frac{d}{dt} x_1(t) &=& p_1 x_1(t) - p_2 x_1(t) x_2(t)\\ -->
<!--     \frac{d}{dt} x_2(t) &=& p_3 x_1(t) x_2(t) - p_4 x_2(t) - u(t) x_2(t) -->
<!-- \end{eqnarray*} </p> -->

<!-- <p> Here $u(t)$, called an _input signal_ is a function that represents the effort exerted in removal of predators. With this input signal in place, we can now use the model to explore the consequences of a range of removal schedules. In particular, optimization can be used to identify the removal strategy that best achieves some performance goal. </p> -->

<!-- <p> As a concrete example, consider the goal of maximizing harvest over some fixed time period (e.g.~a year). Let's consider two cases. To begin, suppose that the harvesting rate must be fixed throughout the system behaviour. In this case, the harvesting rate is constant. Let's call it $u(t) = u_0$. We can then account for the total harvest by introducing a new state variable that tracks the progress of the harvest: -->
<!-- \begin{eqnarray*} -->
<!--     \frac{d}{dt} x_1(t) &=& p_1 x_1(t) - p_2 x_1(t) x_2(t)\\ -->
<!--     \frac{d}{dt} x_2(t) &=& p_3 x_1(t) x_2(t) - p_4 x_2(t) - u(t) x_2(t)\\ -->
<!--     \frac{d}{dt} x_3(t) &=& u(t) x_2(t) -->
<!-- \end{eqnarray*} </p> -->

<!-- <p> Then the value of $x_3(t)$ is the accumulated harvest from time zero to time $t$. If our goal is to maximize the total harvest over a year, we aim to maximize $x_3(1)$ (where $t$ is in units of years) over the choice of the fixed harvesting effort $u_0$. </p> -->
<!-- [Details of implementation; exercise] -->

<!-- <p> Let's now consider the case in which we can change the harvesting effort through the year: we'll allow the input $u(t)$ to be a time-varying function. One would expect we'd be able to achieve a larger harvest in this case (compared with a fixed harvest rate), because there's more freedom in the design of the harvesting strategy. We'll see below that this is indeed true. However, the optimization task as currently posed is problematic. So far, we've been optimizing over functions (SSE, total harvest) that depend on a finite set of numerical values, and so we searched a finite-dimensional space for the extrema. We're now faced with optimizing over *any* function $u(t)$. This is what's technically called an infinite-dimensional search space. The search for extrema in this case has been investigated in the theory of _optimal control_ and closely related work in the _calculus of variations_. See (Lenhart and Workman, 2007) if you'd like to learn more. Here, we'll restrict ourselves to a simple short-cut for solving this problem. We need to find a way to _parameterize_ the set of possible input curves $u(t)$. A common strategy is as follows: -->

<!-- __Sample-and-hold input parametrization__ -->

<!-- <p> We suppose that the input $u(t)$ (harvesting effort) is a piece-wise constant function, which changes values only at pre-specified timepoints (e.g. monthly, as shown in figure XYZ). Then we can specify any such function as a set of 12 numerical values, one for each month: ($u_1$, $u_2$, $u_3$, ..., $u_{11}$, $u_{12}$). As youd expect, the optimization task (which is now an optimal control task) can be solved by searching for the maximizer of total harvest over this 12-dimensional search space. More details in Lin et al. 2014. </p> -->
<!-- [Implementation]; note: easiest way to implement is as a series of separate simulations. Also note:  solution is not differentiable at timepoints where $u$ jumps. -->
<!-- exercise: change number of sub-intervals, note increase/decrease in maximum achieved] -->

## Uncertainty Analysis and Bayesian Calibration

<p> The regression tasks discussed above resulted in estimates of parameter values based on the provided datasets. Regression is usually followed by _uncertainty analysis_, which provides some measure of confidence in those estimated parameter values. For instance, in the case of linear regression, 95\% confidence intervals on the estimates (best fit line slope and intercept) and corresponding confidence intervals on the model predictions are supported by extensive statistical theory, and can be easily generated in R. </p> 

```{r,echo=FALSE}
library(ggplot2)
#data("cars")
```

```{r,echo=FALSE}
#cars.lm <- lm(dist ~ speed, data = cars)
```

```{r,echo=FALSE}
#new.dat <- seq(min(cars$speed),max(cars$speed),by = 0.05)
#pred_interval <- predict(cars.lm, newdata = data.frame(speed=new.dat), 
#                         interval = 'prediction',level = 0.95)
```

The plot below shows 95\% confidence intervals on the model predictions. 

```{r,echo=FALSE}
#plot(cars$speed,cars$dist,xlab="speed", ylab="dist", #main="Regression",pch=16)
#abline(cars.lm,col="firebrick",lwd=2)
#lines(new.dat, pred_interval[,2], col="steelblue", lty=2,lwd=2)
#lines(new.dat, pred_interval[,3], col="steelblue", lty=2,lwd=2)
#legend(4, 105, legend=c("Best-fitted Line", "Prediction interval"),
#       col=c("firebrick", "steelblue"), lty=1:2, cex=0.8)
```

The confint() function finds 95\% confidence intervals on the estimates (best fit line slope and intercept).

```{r}
#confint(cars.lm)
```

This can also be derived using the confidence interval formula: $\text{estimate}\pm \text{critical value}\times\text{standard error}$

```{r}
#lm_summ <- summary(cars.lm)
#c("lower" = lm_summ$coef[2,1] - qt(0.975, df = lm_summ$df[2]) * #lm_summ$coef[2, 2],
#  "upper" = lm_summ$coef[2,1] + qt(0.975, df = lm_summ$df[2]) * #lm_summ$coef[2, 2])
#c("lower" = lm_summ$coef[1,1] - qt(0.975, df = lm_summ$df[2]) * #lm_summ$coef[1, 2],
#  "upper" = lm_summ$coef[1,1] + qt(0.975, df = lm_summ$df[2]) * #lm_summ$coef[1, 2])
```

<p> For nonlinear regression (including calibration of dynamic models), the theory is less helpful, but estimates of uncertainty intervals can be achieved (as discussed below in the section on optimal experimental design). </p>

```{r}
confint(model.nls)
```

```{r}
lm_summ <- summary(model.nls)
c("lower" = lm_summ$coef[2,1] - qt(0.975, df = lm_summ$df[2]) * lm_summ$coef[2, 2],
  "upper" = lm_summ$coef[2,1] + qt(0.975, df = lm_summ$df[2]) * lm_summ$coef[2, 2])
c("lower" = lm_summ$coef[1,1] - qt(0.975, df = lm_summ$df[2]) * lm_summ$coef[1, 2],
  "upper" = lm_summ$coef[1,1] + qt(0.975, df = lm_summ$df[2]) * lm_summ$coef[1, 2])
```

<p> __Bayesian methods__ address the regression task by combining calibration and uncertainty in a single process. The basic idea behind Bayesian analysis (founded on Bayes Theorem, which may be familiar from elementary probability), is to start with an initial parameter estimate (analogous to the initial guess needed in nonlinear regression) and then use the available data to refine that estimate. The difference is that instead of the initial guess and refined estimate being single numerical values, they are _distributions_. In Bayesian terminology, we being with a _prior_ distribution, which may be informed by expert knowledge (e.g.~a normal distribution centered at a good initial estimate), or may be a wild guess (e.g.~a uniform distribution over a wide range of possible values). Application of a Bayesian calibration scheme uses the available data to generate an improved distribution of the parameter values, called the  _posterior_ distribution. A successful Bayesian calibration could take a `wild guess' uniform prior and return a tightly-centered posterior. Uncertainty can then be gleaned directly from the posterior, by reporting, e.g.~ a 95% confidence interval. One commonly cited concern with Bayesian techniques is the dependence of the result on the rather subjective selection of a prior. This is analogous to concerns about any bias introduced by choice of initial guess in the non-Bayesian analysis described above. In both cases, multi-start techniques can be used to address this issue. </p>

<p> Here we'll consider a simple numerical implementation of a Bayesian approach: Approximate Bayesian Computation (ABC). This approach is based on a simple idea: the _rejection method_, in which we sample repeatedly from the prior distribution and reject all samples that do not satisfy a pre-specified tolerance for quality of fit. (This is reminiscent of the selection step in genetic algorithms: culling unfit members of a population.) </p>

For the Michaelis-Menten example, the estimated values for $K$ and $V_m$ are 0.4398016 and 0.0054252 respectively. Now we simulate numbers for $K$ and $V_m$, and accept the value if the distances between the simulated values and the true values are under the threshold.

```{r}
# use the original dataset
S = c(3.6, 1.8, 0.9, 0.45, 0.225, 0.1125, 3.6, 1.8, 0.9, 0.45, 0.225, 0.1125, 
      3.6, 1.8, 0.9, 0.45, 0.225, 0.1125, 0)
v = c(0.004407692, 0.004192308, 0.003553846, 0.002576923, 0.001661538, 
      0.001064286, 0.004835714, 0.004671429, 0.0039, 0.002857143,
      0.00175, 0.001057143, 0.004907143, 0.004521429, 0.00375, 0.002764286,
      0.001857143, 0.001121429, 0)

# functions to randomly draw values for K and Vm
draw_K <- function () {
  return (runif(1, min=0, max=1))
}
draw_Vm <- function () {
  return (runif(1, min=0, max=0.01))
}
```

```{r}
# simulate values for v
simulate_data <- function (S, K, Vm) { 
  return(Vm*S/(K+S))
}

# calculate distances between true and simulated values
compare_with_squared_distance <- function (true, simulated) {
  distance = sqrt(sum(mapply(function(x,y) (x-y)^2, true, simulated)))
  return(distance)
}

# returns true if the distance is less than the given threshold, false otherwise
accept_or_reject_with_squared_distance <- function (true, simulated, acceptance_threshold) {
  distance = compare_with_squared_distance(true, simulated)
  if((distance < acceptance_threshold) ) return(T) else return(F)
}

# returns the sampled K and Vm values and if they are rejected or not
sample_by_rejection <- function (true_data, n_iterations, acceptance_threshold, accept_or_reject_function) {
  number_of_data_points = length(true_data)
  accepted_or_rejected <- vector(length = n_iterations)
  sampled_K <- vector(length = n_iterations, mode = "numeric")
  sampled_Vm <- vector (length = n_iterations, mode = "numeric")
  for (i in 1:n_iterations){
    K <- draw_K()
    Vm <- draw_Vm()
    parameters = list("K"=K, "Vm"=Vm)
    simulated_data <- simulate_data(S,K,Vm)
    accepted_or_rejected[i] = accept_or_reject_function(true_data, simulated_data, acceptance_threshold)
    sampled_K[i] = K
    sampled_Vm[i] = Vm
  }
  return(data.frame(cbind("accepted_or_rejected" = accepted_or_rejected, "sampled_Ks" = sampled_K, "sampled_Vms" = sampled_Vm)))
}
```

Here we randomly sample $K$ and $V_m$ 200000 times, it can be seen from the histograms that the values that occur most frequently are close to the estimated values derived from the non linear regression model.

```{r}
# set seed
set.seed(132)
# simulate 200000 times with a threshold of 0.001
sampled_parameter_values_squared_distances = sample_by_rejection(v, 200000, 0.001, accept_or_reject_with_squared_distance)

# number of accepted values among 200000 simulations
sum(sampled_parameter_values_squared_distances$accepted_or_rejected)
```
```{r}
c1 <- rgb(173,216,230,max = 255, alpha = 80, names = "lt.blue")
c2 <- rgb(255,192,203, max = 255, alpha = 80, names = "lt.pink")
prior1 <- hist(sampled_parameter_values_squared_distances[1:3000,3],col = c1,
               xlab = "K", main="Prior Versus Posterior Distribution for K") 
post1 <- hist(sampled_parameter_values_squared_distances[which(sampled_parameter_values_squared_distances$accepted_or_rejected==1),3], col=c2, xlab = "K", add=TRUE)
abline(v=0.005425,col="firebrick",lwd=2)
legend("bottomright",legend=c("Prior Distribution for K", "Posterior Distribution for K"),
       inset=.02, fill=c(c1,c2), cex=0.8)

prior2 = hist(sampled_parameter_values_squared_distances[1:2500,2],col = c1,
              xlab = "Vm", main="Prior Distribution for Vm") 
post2 = hist(sampled_parameter_values_squared_distances[which(sampled_parameter_values_squared_distances$accepted_or_rejected==1),2],col = c2, xlab = "Vm", add=TRUE)
abline(v=0.439802,col="firebrick",lwd=2)
legend("bottomright",legend=c("Prior Distribution for Vm", "Posterior Distribution for Vm"),
       inset=.02, fill=c(c1,c2), cex=0.8)
```

<p> Typically, approximate Bayesian computation is implemented as an iterative method, in which a sequence of rejection steps is applied, withe the prior being refined at each step (generating, in essence, a sequence of posterior distributions, the last of which is considered to be the best description of the desired parameter estimates). </p>

```{r,message=FALSE,warning=FALSE}
library(EasyABC)
```

```{r}
# set seed
set.seed(123)

# simulate data
mm_model <- function(par) {
  Vm <- par[1]
  K <- par[2]
  S <- S
  samples <- S*Vm/(S+K)
  return(samples)
}

# set tolerance levels
tolerance=c(8,3)

# identify prior distribution for parameters
mm_prior = list(c("unif",0,0.01),c("unif",0,1))
```

```{r}
# use ABC_sequential() to obtain a rough posterior distribution of parameter values
ABC_sequential <- ABC_sequential(method="Beaumont", model=mm_model,prior=mm_prior, 
                                 nb_simul=20000, summary_stat_target=v,
                                 tolerance_tab=tolerance)
```

After 20000 simulations, it can be seen from the histograms that the most frequent values are close to the estimated values derived from the non linear regression model.

```{r}
c1 <- rgb(173,216,230,max = 255, alpha = 80, names = "lt.blue")
c2 <- rgb(255,192,203, max = 255, alpha = 80, names = "lt.pink")

hist(runif(15000,0,0.01),main="Prior Versus Posterior Distribution for K",
     col = c1,xlab = "K")
hist(ABC_sequential$param[5000:10000,1],col = c2, add=TRUE)
abline(v=0.005425,col="firebrick",lwd=2)
legend("topleft",legend=c("Prior Distribution for K", "Posterior Distribution for K"),
       inset=.02, fill=c(c1,c2), cex=0.8)

hist(runif(10000,0,1),main="Prior Versus Posterior Distribution for Vm",
     col = c1,xlab = "K")
hist(ABC_sequential$param[5000:10000,2],col = c2, add=TRUE)
abline(v=0.439802,col="firebrick",lwd=2)
legend("topleft",legend=c("Prior Distribution for Vm", "Posterior Distribution for Vm"),
       inset=.02, fill=c(c1,c2), cex=0.8)
```

<!-- ### Optimal Experimental Design  -->

<!-- <p> Experimental design involves choices about what experiments to perform, what measurements to take, and, for time-varying processes, when to take those measurements. Most experimental design exercises are carried out in the face of constraints on available resources (time, personnel, reagents, funding, etc.). The goal of experimental design is then to identify the 'best' experiments to execute, given the imposed resource constraints. In cases where we can express the 'quality' experimental results in terms of a numerical performance, and can likewsise express constraints numerically, we can then frame the experimental design task as an optimization task; this is _optimal experimental design_. There is a rich history of optimal experimental design for linear systems (Pukelsheim, 2006.). Here, we consider a more challenging task of optimal experimental design in a nonlinear setting. Consider the case in which a nonlinear regression has been applied to find a best fit model to some pre-existing data, along with an estimate of the quality of that fit. We then consider an experimental design task that targets the refinement of that estimate: the `best' experimental results will be those that most shrink the confidence intervals when that data is included in the regression analysis. </p> -->

<!-- <p> We quantify the degree of confidence in the parameter estimate with the _Fisher Information Matrix_ (FIM) which is constructed from (i) the sensitivities of the model predictions to the values of the model parameters, and (ii) the variance in the data. The sensitivities account for how much effect a changing the value of model parameters affects the model predictions that will be used for calibration. If the parameter values have little influence over those predictions, then we can't expect to have much confidence in estimate the parameter values based on tuning of those predictions. Likewise, if the dataset shows wide variation, then we'll expect to have less confidence from tuning the model to agree with the observed behaviours. A simple measure of overall confidence is provided by the determinant of the Fisher information matrix. (This is referred to as D-optimality in a common alphabetic classification of related optimality objectives). This determinant is a estimate of the size of a 95% confidence ellipsoid that corresponds to a level set of the cost surface (Figure XYZ). </p> -->
<!-- [IMPLEMENT D-optimal design for simple model, e.g. Michaelis-Menton or Hill.] -->

<!-- ### Optimal distribution of intracellular metabolic fluxes (Flux Balance Analysis) -->

<!-- <p> We've been focusing on the use of iterative optimization algorithms to solve nonlinear optimization tasks. For some problems, such sophisticated (and computationally intensive) techniques are not need. One important class of such tasks is referred to as _linear programming_, in which we seek to find the optima of a linear objective function while in as search space that is bounded by linear constraints. An example: -->
<!--  \begin{eqnarray*} -->
<!--  \mbox{maximize} \ 3x_1 - 4x_2 -->
<!--  \mbox{subject to} \ 2x_1 + 5x_2 < 3 \ \mbox{and} \ -6x_1 + 4x_2 < -6 -->
<!--  \end{eqnarray*} </p> -->
<!-- <p> There are iterative algorithms for solving such problems, that are guaranteed to reach the solution after a finite number of steps (in contrast to the global optimization algorithms discussed earlier). These algorithms, such as the simplex method, step from vertex to vertex within the valid solution space until they arrive at the desired extremum. [Fig XYZ] </p> -->
<!-- [Implement solution to toy problem.] -->

<!-- <p> Linear programming task arise regularly in planning, scheduling, and other management tasks. One biological context in which they occur is prediction of  the distribution of intracellular metabolic fluxes. Genomic analysis allows us to identify the suite of metabolic enzyme that can be expressed within a given cell, and thus identify the set of metabolic processes that may be active in that cell. If we think of these metabolic processes occurring in the context of balanced cell growth, we can safely assume that the net production rate of all metabolic intermediates is zero (i.e.~rate of production is balanced by rate of consumption). If we assign a label to the rate of each reaction (i.e.~the reaction flux) in an intracellular metabolic network, this balance condition imposes a set of linear constraints on the components of a flux vector. For example, consider an idealized cell with a simple metabolism as in figiure XYZ. The balance conditions at intermediate metabolites $s_1$, $s_2$, $s_3$, and $s_4$ gives rise to constraints: -->
<!--  $v_1=v_3$, $v_2=v_4$, $v_3+v_4=v_5$, $v_5=v_6+v_7$. Next, consider the case in which the rates of uptake and secretion of (at least some) nutrients and products have been measured. In our case, suppose we have measured $v_1=1 uM/hr$, $v_2=3 um/hr$. .... Not surprisingly, these few measurements of fluxes outside the cell do not provide enough information to determine the values of all fluxes in the intracellular metabolic network. (This is described by saying that the system is _underdetermined_.) Nevertheless, there are cases in which we want to make such a prediction. To address such cases, the technique of Flux Balance Analysis imposes an assumption about the cell's internal flux allocation. The assumption is that the cell regulates the internal fluxes to maximize some objective. In the case of microbial cells, this is most commonly presumed to be growth rate. With this objective in place, expressed typically in terms of maximizing biomass: written as a linear combination of particular intracellular reactions, the task of predicting all intracellular fluxes becomes a linear programming task. </p> -->
<!--  [implement example] -->

## References

Ashyraliyev, M., Fomekong‐Nanfack, Y., Kaandorp, J. A., & Blom, J. G. (2009). Systems biology: parameter estimation for biochemical models. The FEBS journal, 276(4), 886-902.
 
 <!-- Lenhart, Suzanne, and John T. Workman. Optimal control applied to biological models. CRC press, 2007. -->

 <!-- Lin, Qun, Ryan Loxton, and Kok Lay Teo. "The control parameterization method for nonlinear optimal control: a survey." Journal of Industrial and management optimization 10.1 (2014): 275-309. -->

 Beaumont, Mark A. "Approximate bayesian computation." Annual review of statistics and its application 6 (2019): 379-403.
 
 Cho, Yong-Soon, and Hyeong-Seok Lim. "Comparison of various estimation methods for the parameters of Michaelis-Menten equation based on in vitro elimination kinetic simulation data." Translational and clinical pharmacology 26.1 (2018): 39-47.
 
 Fairway, Julien, (2002) Practical Regression and Anova using R. https://cran.r-project.org/doc/contrib/Faraway-PRA.pdf
 
 <!-- Pukelsheim, Friedrich. Optimal design of experiments. Society for Industrial and Applied Mathematics, 2006. -->




